---
title: "Homework 1"
author: "Ali Early"
output: github_document
---

## Problem 0
As evidenced throughout this assignment.



## Problem 1 
`library(readr)`

```{r}
nyc_transit_data = 
  read_csv("data:/NYC_Transit.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    starts_with("route"),
    entry,
    vending,
    entrance_type,
    ada
    ) |>
    mutate(
      across(starts_with("route"), as.character),
      entry = case_match(
        entry,
        "YES" ~ TRUE,
        "NO" ~ FALSE
      )
    )
```

### Paragraph on the process
This dataset presents findings related to NYC subway entry and exit points, along with other critical pieces of information about various subway station that a traveler may find useful. Some variables including the subway line, station details (e.g., name, lat, long.), routes, entrance type (e.g., stairs), entry and exit availability, vending options, and ADA compliance. Some of the variables are characters (e.g., station name, line), others are numerical (e.g., lat., long.), and others are logical (e.g., ADA). Additionally, the overall dataset has 1,868 rows and 19 columns. My cleaning process involved using the `janitor` code along with "select" to shift the order of the columns for ease of viewing. My cleaning process also involved using `janitor` to adjust column names, `select()` to keep relevant variables for the homework, and change entry from "YES"/"NO" into TRUE/FALSE. The data aren't completely tidy yet because the routes are spread out across columns. 

### Questions answered by the dataset
1. How many distinct stations are there? 
```{r}
distinct_stations = 
  nyc_transit_data |>
  distinct(line, station_name, .keep_all = TRUE)
nrow(distinct_stations)
```


2. How many stations are ADA compliant?
```{r}
distinct_stations |>
  filter(ada == TRUE) |>
  nrow()
```


3. What proportion of station entrances / exits without vending allow entrance?
```{r}
nyc_transit_data |>
  filter(vending == "NO") |>
  summarize(prop_entry = mean(entry))
```


4. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r}
nyc_transit_tidy = 
  nyc_transit_data |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )

a_train_stations = 
  nyc_transit_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name, .keep_all = TRUE)

nrow(a_train_stations)

a_train_stations |>
  filter(ada == TRUE) |>
  nrow()
```



## Problem 2
`library(readxl)`

```{r}
mr_trash_wheel =
  read_excel("data:/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1
  ) |> 
  janitor::clean_names() |> 
  select(-starts_with("unnamed")) |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel"
  )


prof_trash_wheel = 
  read_excel(
    "data:/202610 Trash Wheel Collection Data.xlsx", 
    sheet = "Professor Trash Wheel", 
    skip = 1
  ) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.numeric(year),
    plastic_bottles = suppressWarnings(as.numeric(plastic_bottles)),
    trash_wheel = "Professor Trash Wheel"
  )


gwynnda_trash_wheel = 
  read_excel(
    "data:/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    skip = 1
  ) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.numeric(year),
    trash_wheel = "Gwynnda"
  )

trash_wheel_tidy = 
  bind_rows(mr_trash_wheel, prof_trash_wheel, gwynnda_trash_wheel)
```

### Paragraph and inline R
The dataset presents `trash_wheel_tidy` with 1,269 observations (i.e., dumpster numbers). Some columns across the three sheets of interest ("Mr. Trash Wheel," "Professor Trash Wheel," and "Gwynnda the Good Wheel") include columns related to the Healthy Harbor initiative, such as `dumpster`, `month`, `year`, `weight`, `volume`, `plastic`, `polystyrene`, `cigarettes`, `glass`, `plastic`, `wrappers`, `sports`, and `homes`. Most of these variables are numeric but not all such as month, year, and date of data collection. 

```{r}
total_prof_weight = 
  trash_wheel_tidy |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  summarize(total_weight = sum(weight_tons, na.rm = TRUE)) |> 
  pull(total_weight)


gwynnda_june_2022_cig_butts = 
  trash_wheel_tidy |> 
  filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |> 
  summarize(total_butts = sum(cigarette_butts, na.rm = TRUE)) |> 
  pull(total_butts)
```

1. What was the total weight of trash collected by Professor Trash Wheel? 
294.96 tons


2. What was the total number of cigarette butts collected by Gwynnda in June 2022?
18,120 butts


## Problem 3
```{r}
mci_baseline = 
  read_csv("data:/MCI_baseline.csv", skip = 1) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    ),
    age_at_onset = suppressWarnings(as.numeric(age_at_onset))
  ) |> 
  filter(is.na(age_at_onset) | age_at_onset > current_age)



mci_amyloid = 
  read_csv("data:/mci_amyloid.csv", skip = 1) |> 
  janitor::clean_names() |> 
  rename(id = study_id) |> 
  pivot_longer(
    cols = baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |> 
  filter(!is.na(amyloid_ratio))



only_in_baseline = anti_join(mci_baseline, mci_amyloid, by = "id")
only_in_amyloid = anti_join(mci_amyloid, mci_baseline, by = "id")


mci_combined = inner_join(mci_baseline, mci_amyloid, by = "id")

write_csv(mci_combined, "data:/mci_combined.csv")
```
### Discuss important steps 
1. Import process
I first downloaded the datasets then imported it using the `read_csv` code to read the files. I also used the `janitor` code like the last problem to clean the data. Next, I made better/cleaner variables of sex (male-female) and carriers (carriers-non-carriers). As for relevant features, both have quite a bit of missing data which is represented by NA or blanks. 

2. How many participants were recruited, and of these how many develop MCI?
```{r}
n_recruited = nrow(read_csv("data:/MCI_baseline.csv", skip = 1))

n_developed_mci = 
  mci_baseline |> 
  filter(!is.na(age_at_onset)) |> 
  nrow()
```
483 total, 479 left


3. What is the average baseline age? 
```{r}
avg_baseline_age = 
  mci_baseline |> 
  summarize(mean_age = mean(current_age, na.rm = TRUE)) |> 
  pull(mean_age)
```

65.03 years old


4. What proportion of women in the study are APOE4 carriers?
```{r}
prop_female_carriers = 
  mci_baseline |> 
  filter(sex == "female") |> 
  summarize(prop = mean(apoe4 == "carrier")) |> 
  pull(prop)
```
0.3 or 30% are women in the study are APOE4 carriers.


```{r}
mci_amyloid = 
  read_csv("data/mci_amyloid.csv", skip = 1) |> 
  janitor::clean_names() |> 
  rename(id = study_id) |> 
  pivot_longer(
    cols = baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |> 
  filter(!is.na(amyloid_ratio))


only_in_baseline = anti_join(mci_baseline, mci_amyloid, by = "id")
only_in_amyloid = anti_join(mci_amyloid, mci_baseline, by = "id")


mci_combined = inner_join(mci_baseline, mci_amyloid, by = "id")


write_csv(mci_combined, "data:/mci_combined.csv")
```

### Discuss important steps 
1. Import process
I first downloaded the datasets then imported them using the `read_csv` code with `skip = 1` to read the files properly. I also used `janitor` to clean the column names, and renamed study_id to id. Next, I used `pivot_longer` on `cols = baseline:time_8` to tidy the longitudinal data into `visit` and `amyloid_ratio` columns, and used `filter` to ditch missing values. As for relevant features, both datasets have missing data represented by NAs or blanks. I checked for unmatched participants using `anti_join` before combining them with `inner_join`.


2. The baseline dataset has 479 participants with demographic variables (e.g., current_age, sex, education, apoe4, age_at_onset). The longitudinal amyloid dataset has repeated measurements of the amyloid ratio for five study visits. Merging both datasets has a combined dataset of 2,196 observations across 471 participants, linking baseline demographics to individual biomarkers.


3. Comparing ids using `anti_join()` shows that eight participants were only in the baseline dataset. This could be due to loss to follow-up. Also, 16 participants were only in the amyloid dataset.Combining the datasets `(inner_join)` maintains the 471 participants who are in both files. The total dataset has 2,196 total observations.

`write_csv(mci_combined, "data:/mci_combined.csv")`
